Busch, Khakhalin. Intrinsic temporal tuning

### 2019-02-02 reviews and responses.md

We think that this word well reflects the nature of the property we are describing (the property of being easily excitable, and able to generate spikes), and we also think that we define it rather precisely within thiss paper. We prefer it to other alternatives, as they are rather too general ("intrinsic excitability") or too bulky ("ability to produce spikes in response to stimulation").

Of the two options offered by the reviewer, the first one ("Changes in f-I curve") is both too long, and does not reflect the nature of our measure, as with it we are trying to assess average ability to produce spikes, not a change of any sort. The second one sounds good and standard, but is actually somewhat misleading, as the term "Avereage firing rate" is defined rather precisely as the average rate at which a neuron fires in the network; usually either in the context of in-vivo recordings, or in-vitro spontaneous activity. Accordingly, a true "Average firing rate" for our neurons would depend both on their intrinsic properties (perhaps the measure we call "spikness here"), *and* the strength of their synaptic inputs, as well as their position within the network. In short, while "Average firing rate" is an important property, in this study we did not address it, nor had an opportunity to observe a proxy for it.

> Page 3: "traces left by multi sensory stimuli in the tectum of freely behaving tadpoles" -> unclear what is meant by the this sentence until later; suggest re-word

Reworded: "for the first time we were able to look at tectal network retuning in response to multisensory stimuli in freely behaving tadpoles."

@@

> Results

> Page 4, line 6 "the stimuli we used were weaker" What is meant, the stimuli in the previous studies were weaker, or the stimuli in the present study? Maybe say the previous stimuli were very high contrast and did not reflect contrasts used in behavioral studies. Here, we presented a checkerboard"...

Changed to "the stimuli we used in the present study were weaker".

@@

> Page 4 line 19 "spiky" again, imprecise

We like the word "spiky", as it captures the phenomelogical effect without making assuptions about underlying mechanisms. It is also intuitive and short, which aids understanding.

@@

> Page 4, lines 21-24. I'm really confused about the degrees of freedom. Are you all loading in all observations from all cells? If so, the assumption of the statistical test that these are independent observations is violated. Many of these values were measured in the same cells with different current levels, so they are clearly not independent measurements. The analysis should essentially contain one observation per cell. (One "gain" value or number of spikes to some stimulus, etc..) Or could do a multi-variate ANOVA...

We did in fact run a linear model with repeated measures that may be vaguely described as multi-variate ANOVA; see responses above.

> Page 4: Lines 27, 28: This sentence is missing conjunctions "in order to map their amplitude transfer function" ?

Reworded: "We then mapped the amplitude tuning of neurons (also known as amplitude transfer function, or gain), by looking at how an increase in transmembrane conductance translated into increased spike output."

@@

> Page 4: Lines 27, 28: What exactly is meant by the amplitude transfer function or amplitude tuning? Is there an equation?

At this point we describe tuning as an empirical curve (something that can be plotted in a figure, and compared from cell to cell), but do not yet use any equations. Later in the text we parameterize these curves, and provide all related equations.

@@

> Page 4, lines 27 - 33: what figure panels are being described here?

We added a reference to (Fig 1D,E).

> Page 6, lines 10-20. The authors point out that the amount of response adaptation in the conductance vs. number of spike curves varies as a function of the stimulus that the animal was exposed to. While it is true that this information can be gleaned by examining Figure 1E, the authors would be well served to make a second index, such as an adaptation index, in a panel F. This would really drive home the points the authors would like the reader to take away from the Figure. (I missed it the first time through, I just saw the reduced gain in Fig. 1E, not the adaptation.)

We introduce adaptation index (parametrized quantification of temporal tuning) later in the text.

@@

> Page 6, lines 10-20. Need to cite Figure 1E

Done.

> Page 6, line 22: "homologous to superior colliculus"

> Page 7, line 3: after multi sensory stimulation, "tectal neurons were more excitable than after visual stimulation alone). Many issues. The average number of spikes is reported.. Is this across all conductance curve conditions or just one of them?

The averages were calculated across all testing conditions; we added this info to the sentence: "after four hours of multisensory stimulation, tectal neurons were more excitable than after visual stimulation alone (0.6$\pm$0.4 spikes, F(1,689)=11.2, p=8e-4; and 0.7$\pm$0.6 spikes, F(1,665)=41.7, p=2e-10, for sync and async respectively, across all testing conditions)."

@@

The averages were calculated across all testing conditions; we added this info to the sentence: "after four hours of multisensory stimulation, tectal neurons were more excitable than after visual stimulation alone (0.6$\pm$0.4 spikes, F(1,689)=11.2, p=8e-4; and 0.7$\pm$0.6 spikes, F(1,665)=41.7, p=2e-10, for sync and async respectively, across all testing conditions)."

> Next, in Fig 1E, "sync" and "flash" and "looming" all look pretty similar to each other.

We think that it rather reasonable that effects of going through some stimulation, compared to control, are stonger, than differences between different types of stimulation. We explore this question in detail further in the text.

@@

> Figure 3E and G: how can we reconcile these accounts? In each of the groups (3G), there is at best a very weak relationship between temporal tuning and synaptic duration. Somehow, if you take the means of all of these groups, you can fit a line through them (Figure 3E). Why is that the right thing to do? Aren't the field of individual points (partially-transparent background) and the data plotted in 3G the relative quantities?

We now added a paragraph describing this situation in detail, and reference a textbook name for this type of situation, when between-group and within-group analyses yield seemily opposite results (Simpson's paradox).

> Figure 3C shows that monosynaptic amplitude and synaptic duration are correlated, as one would expect from the definition of synaptic duration (center of mass). Basically, as the monosynaptic amplitude gets bigger, the center of mass will shift earlier and earlier. So what independent information is really being conveyed in the synaptic duration? The correlation in 3C is so strong, I don't see that it is truly an independent quantity that teaches us something beyond the monosynaptic input strength. The monosynaptic input strength is much more defined, why not use that for examining correlation with temporal tuning (or adaptation index), average firing rate, and gain. The "synaptic duration" isn't really a measure of synaptic duration, which makes Page 10, lines 3-23 really hard to interpret.

When comparing temporal properties of synaptic inputs to that of intrinsic tuning, it is important to make the two measures as similar (matching) as possible, as methodologically it is supposed to increase the power of our comparisons. In practice, if for intrinsic properties we quantify non-linearity of responses to currents of different duration, then we need to estimate current duration for our synaptic inputs. In other words, while synaptic current duration is largerly just a different way to quantify the relative impact of polysynaptic activity, it is a necessary step for a good analysis. We now reworded two sentences at the beginning of paragraph in question, to hopefully make this point more clear:

"To better match and compare temporal properties of synaptic inputs to those of intrinsic tuning, we calculated average "synaptic current duration" for each cell as a temporal "center of mass" of currents within the first 700 ms after optic chiasm stimulation (see "Methods"). Neurons with different contribution of early and late synaptic responses naturally had different synaptic current duration: cells with strong monosynaptic inputs had shorter currents, while polysynaptic activity made synaptic currents longer (Figure 3C; p=2e-16, r=$-$0.78, n=168)."

@@

> Figure 3G, page 10 lines 24-40. Creation of "super groups" is totally bogus. This part should be dropped. I see no strong significant relationship between temporal tuning and synaptic duration, and am not convinced synaptic duration is a meaningful quantity.

We now removed this paragraph. In response to this comment, we initially attempted a different analysis of the interaction between synaptic and intrinsic properties across groups, but while some reasonable types of comparisons were significant, this analysis lacks power, and relies too much on post-hoc reasoning, so we decided to follow the advice and just remove this paragraph entirely. While we still believe that stronger stimulation (in terms of its effect on average intrinsic properties) generally caused stronger disruption of the synaptic-intrinsic correlation, this statement is hard to fully justify with our current data.

### 2019-02-02 main.tex
tex: lots of tiny edits + extra ref

% TODO LIST

% In Discussion, mention that studying post-hyperpolarization currents would be a great next step

% Intro: be more specific that we're not talking about STG or spinal (with refs). That in FF intrinsic is often underplayed, and that the same is tru for neuro-inspired machine learning. Check my own response to elife rejection, draw from there.

% Explain why homeostatic plasticity in a feature space makes a network more sensitive to unusual stimuli

It is often assumed, in fields as diverse as connectomics and machine learning, that the main difference between functional and dysfunctional neural networks lies in their connectivity \citep{takemura2014,hildebrand2017,bassett2017network,reimann2017}. Biological neurons, however, are tuned in ways that go well beyond adjusting one “strength” value per synapse: cells within the same network often demonstrate variability of activation thresholds \citep{kole2012}, production of bursts \citep{popovic2011}, inactivation by strong inputs \citep{bianchi2012}, and more. This diversity of tuning relies on coordinated changes of many parameters across a complex, multivariate landscape \citep{oleary2013}, as neuronal phenotypes are shaped by sensory experiences, and adjusted by modulatory inputs \citep{evans2015}. The dysregulation of intrinsic plasticity affects network dynamics \citep{tien2018}, and can lead to a loss of function \citep{marcelin2009}. And yet, with an obvious exception of oscillatory networks \citep{marder2011,picton2018control}, for many brain areas it is still unclear whether variation of intrinsic phenotypes serves as a defining aspect of network topology and architecture, or whether it is just a consequence of transfer function normalization \citep{titley2017}.

The optic tectum of the \textit{Xenopus} tadpole is an ideal model for exploring these questions: it is a highly malleable distributed network of about $10^4$ neurons \citep{pratt2013}, involved in stimulus discrimination and sensorimotor transformations \citep{dong2009,khakhalin2014}. In development, tectal neurons acquire diverse intrinsic and synaptic phenotypes, that are also shaped by sensory experiences \citep{xu2011,ciarleglio2015}. Circuits in the tectum can learn and reproduce the temporal dynamics of inputs to which they were exposed \citep{pratt2008}: a property that could in principle be achieved through synaptic changes alone \citep{lukovsevivcius2009}, but which is more likely to involve intrinsic temporal tuning \citep{narayanan2008,beatty2014}. Finally, tectal neurons exhibit strong Na channel inactivation, which seems to play a role in collision detection \citep{jang2016}, and is one of the targets for intrinsic plasticity \citep{bianchi2012}.

In this study, we asked two specific questions about the properties of intrinsic plasticity in tectal networks. First, we asked whether intrinsic plasticity is limited to changes in excitability, or whether it is more nuanced and can differentially adjust neuronal responsiveness to inputs with different dynamics. Second, we checked whether changes in intrinsic properties in the tectum are a homeostatic response to each cell's history of synaptic activation, or if they are independent of synaptic properties. In our previous large-scale census of tectal cells \citep{ciarleglio2015}, we observed no interaction between intrinsic and synaptic phenotypes, and also, despite an extensive search, we did not detect signs of temporal tuning (ibid,  figures 2, 4). However, we propose that the standard current-clamp protocols used in previous studies \citep{pratt2007,hamodi2014} were not adequate to detect important changes in the function of voltage-gated channels (see "Discussion").

Here we show that the intrinsic plasticity of tectal neurons goes beyond changes in average spikiness, and supports temporal selectivity that can be reshaped by sensory experience. Furthermore, we show that the tuning of intrinsic properties is coordinated with the duration of synaptic inputs received by each cell. These results rely on two methodological innovations. First, instead of using current injections, we employed the dynamic clamp technique, which allowed a more realistic simulation of synaptic conductances \citep{prinz2004}. Unlike for more common voltage and current clamp techniques, in dynamic clamp, the electric current injected into the cell is dynamically adjusted, based on a predefined formula that depends on cell membrane potential and time. And second, instead of relying on one type of sensory stimulation, we used five different stimulation protocols, and compared their effects on intrinsic tuning. The use of different sensory modalities also gave us insight into an unrelated, but equally intriguing question of multisensory integration in the brain \citep{deeg2009,felch2016,truszkowski2017}, as for the first time we were able to look at the traces left by multisensory stimuli in the tectum of freely behaving tadpoles.

@@

All analysis scripts, and summary data for every cell, can be found at:  \url{https://github.com/khakhalin/Dynamic-clamp-2018}

@@

Our first question was whether our stimulation protocols caused changes in intrinsic excitability of tectal neurons. From previous studies, we knew that in tadpoles exposed to four hours of LED flashes tectal neurons became more excitable \citep{aizenman2003,ciarleglio2015}. However, the stimuli we used were weaker, and similar to those used in behavioral experiments \citep{khakhalin2014,james2015,truszkowski2017}. We presented a checkerboard pattern that inverted once a second either instantaneously (Figure 1C left, dubbed “Flash”), or with a slow transition over the course of a second (old black squares shrank to white, while new black squares grew from old white squares; Figure 1C right; dubbed “Looming”; see “Methods”).

\begin{figure*}
\includegraphics[width=\linewidth]{fig1.pdf}
\caption{
Overview of experimental design, and summary of dynamic clamp results. (\textbf{A}). Positions of tectal neurons that were recorded. (\textbf{B}). Sample data from a dynamic clamp experiment. Bottom row: the dynamics of conductances G(t) of four different durations simulated by the dynamic clamp system. Middle row: the currents I(t) dynamically injected into a cell based on conductances of 4 different durations and 3 different amplitudes. Top row: resulting voltage traces V(t) that were recorded and analyzed. (\textbf{C}). A schematic of visual conditioning, in “Flash” (left) and “Looming” (right) groups. (\textbf{D}). The number of spikes produced by all neurons in all experiments, split by input peak conductance, and plotted against conductance duration. Black lines show respective averages. (\textbf{E}). A summary of data from (d), presented as averages and 95\% confidence intervals.}
\end{figure*}

After conditioning, we excised the brain, obtained whole cell patch clamp recordings from neurons in the tectum (Figure 1A), and counted spikes produced in dynamic clamp, in response to simulated synaptic conductances of different durations and amplitudes (Figure 1B). We used conductances of 4 different durations (100, 200, 500, and 1000 ms), and 3 different amplitudes (peak conductances of 0.2, 0.5, and 1.0 nS), matching the range of synaptic currents observed in tectal circuits \textit{in vivo} (see "Methods"). Contrary to our expectations, in tadpoles that were exposed to instantaneous checkerboard inversions (flashes), tectal neurons on average became not more, but less spiky, and generated 0.4$\pm$0.4 spikes, across all types of dynamic clamp inputs, compared to 0.9$\pm$1.0 spikes in control (Figure 1D, left column; F(1,677)=30.4, p=5e-8, n cells=28, 29. Here and below, F-values are reported for a multivariate fixed effects analysis of variance with selected interactions, where cell id is used as a repeated measures factor; see "Methods" for a detailed description). Neurons exposed to looming transitions also spiked less than control neurons (0.6$\pm$0.4 spikes; F(1,629)=9.6; p=0.002; n=28, 25), but more than those exposed to instantaneous "flashes" (F(1,641)=3.7; p=4e-6; n=29, 25), which was likewise unexpected. In the tadpole tectum, looming stimuli are known to elicit stronger responses compared to instantaneous inversions \citep{khakhalin2014}, yet the suppression yielded by looming stimuli was weaker.

We then looked at how an increase in transmembrane conductance translated into increased spike output of neurons, to map their amplitude tuning (also known as amplitude transfer function, or gain). Compared to control, neurons from animals exposed to visual stimuli had a flatter amplitude tuning curve, and did not increase their spiking as fast in response to larger conductances (F(1,677)=15.9, p=8e-5; and F(1,629)=8.3, p=0.004 for flash- and looming transitions respectively). The difference in amplitude tuning between these two types of stimulation was not significant (F(1,641)=3.5; p=0.06).

@@

Across input conductances of different lengths (100, 200, 500, and 1000 ms), neurons exposed to flashes and looming stimuli responded differently than in control (F(1, 677)=25.1; p=7e-7; and F(1,629)=12.0; p=6e-4 respectively). In control neurons, longer inputs typically evoked stronger spiking, whereas neurons from stimulated animals had flatter tuning curves, with a plateau, or even a decrease in spike number for longer conductance injections (Figure 1E). In essence, while control neurons “preferred” longer inputs, stimulated neurons developed a preference for shorter synaptic inputs, and this change was more pronounced in neurons exposed to flashes than in those exposed to looming stimuli (F(1,641)=7.5; p=0.006). This suggests that the change in overall intrinsic excitability, and the change in temporal tuning, follow two different kinds of logic. The overall excitability is homeostatic, as neurons became less excitable in response to stronger stimulation. The temporal retuning however can be better described as “adaptive”, as neurons exposed to shorter stimuli (flashes) became relatively \textit{more} responsive to shorter stimuli, and less responsive to longer stimuli, which is opposite of what one would expect for a fully homeostatic retuning. We chose to call this type of plasticity "adaptive", as presumably it means that after exposure to faster stimuli, neurons become more equipped to process faster patterns of activation.

@@

To test this question, we exposed tadpoles to four hours of behaviorally salient “click” sounds \citep{james2015,truszkowski2017}, provided at the same frequency (every second) as visual stimuli in the first set of experiments. We found (Fig 1D, E) that exposure to startle-inducing sounds (group “Sound”) did not lead to significant changes in either average spikiness (0.8$\pm$0.7 spikes; F(1,689)=1.7; p=0.2; n=28, 30), amplitude transfer function (F(1,689)=2.0; p=0.2), or temporal tuning curve (F(1,689)=2.0; p=0.2). This suggests that acoustic stimuli did not strongly activate tectal circuits during conditioning, despite being more behaviorally salient (for our stimuli, at the onset of stimulation, acoustic clicks evoked startle responses in about 50-80\% of cases, compared to 5-10\% for checkerboard inversions \citep{james2015,truszkowski2017}). As one possible explanation, auditory and mechanosensory inputs may have different cellular or subcellular targets in the tectum \citep{bollmann2009}, or they may differently recruit tectal inhibitory circuits \citep{liu2016,hamodi2016}.

We then combined visual and acoustic stimuli in two different ways and looked at the effects of multisensory stimulation on the intrinsic properties of tectal neurons. For some animals, we synchronized the instantaneous checkerboard inversions (flashes) with sound clicks (dubbed “Sync”), while for others we staggered visual and acoustic stimuli by half a period (500 ms; dubbed “Async”). We found (Fig 1D, E) that, after four hours of multisensory stimulation, tectal neurons were more excitable than after visual stimulation alone (0.6$\pm$0.4 spikes, F(1,689)=11.2, p=8e-4; and 0.7$\pm$0.6 spikes, F(1,665)=41.7, p=2e-10, for sync and async respectively). Compared to the “Flash” group, the tuning curves in multisensory groups were less flat, with a stronger effect in the Async group (for amplitude tuning: F(1,689)=1.3, p=0.3, and F(1,665)=2.4, p=0.02 in Sync and Async groups respectively; for temporal tuning: F(1,689)=9.3, p=0.002, and F(1,665)=22.8, p=2e-6 respectively). This suggests that on their own sound clicks had little effect on tectal excitability, but when added to visual flashes, sound clicks negated effects of retuning that visual stimulation would have had (Fig 2C).

\begin{figure*}[!t]
\includegraphics[width=\linewidth]{fig2.pdf}
\caption{
Quantification of changes in temporal tuning in response to sensory experience. (\textbf{A}). An illustration of how “Temporal tuning” and “Amplitude tuning” values were calculated. For the temporal tuning measure, the value of zero corresponds to linear dependency (blue line), positive values - to an accelerating, supralinear curve (red), and negative values - to a plateau-shaped curve (purple, yellow). For amplitude tuning, higher values correspond to faster increase in spiking with increased conductance. (\textbf{B}). Amplitude tuning of neurons across different experimental groups (stars show t-test p<0.05 compared to control). (\textbf{C}). Temporal tuning and average spikiness of neurons in different experimental groups; in each plot all neurons across all groups are shown in gray, while neurons from one target group are shown in color; means are shown as black dots; ellipses represent 95\% normal confidence regions. Two outliers (top right corner) are brought within the axes limits. (\textbf{D}). Same data as in (C), shown as averages for each group, with 95\% confidence intervals. Black arrows show the effects of sound clicks, when they were added to control, and when they were added to “Flashes”, to form two types of multisensory stimuli.}
\end{figure*}

@@

To visualize and interpret differences in neuronal tuning, we quantified each of the three aspects of intrinsic tuning (average spikiness, amplitude tuning, and temporal tuning) with one value per neuron (see Methods). We used the mean number of spikes across all conditions as the measure of "spikiness"; the linear slope of the number of spikes as a function of conductance amplitude as the measure of "amplitude tuning", and the quadratic term of the curvilinear regression for the number of spikes as a function of input duration as the value characterising the "temporal tuning" or adaptation index of each neuron (Figure 2A). The numerical values of these "tuning coefficients" are not easily interpretable, but they capture the character of tuning curves for each neuron (Figure 2A). All three parameters differed across experimental groups: F(5,160)=3.1, p=0.01 for average spikiness (Figure 2C; see "Methods" for model description); F(5,160)=4.8, p=4e-4 for amplitude tuning (Figure 2B); and F(5,160)=3.6, p=4e-3 for temporal tuning (Figure 2C).

@@

Describing neuronal tuning with only few variables allowed us to compare cell-to-cell variability of tuning in different experimental groups. We found that this variability decreased as neurons were modulated away from the baseline (Bartlett test p=2e-9 for amplitude tuning, p=1e-9 for temporal tuning; Figure 2D). Groups that were significantly different in average values also had different variances (F-test with p$<$0.05), such as Control vs. Flash and Control vs. Looming for both amplitude and temporal tuning. This expands on a finding in our previous study \citep{ciarleglio2015} that prolonged patterned stimulation reduces diversity of tuning profiles in the network, reshaping them according to the spatiotemporal characteristics of the stimulus, with stronger stimuli having stronger effects on tuning diversity.

@@

To see whether prolonged sensory stimulation affected synaptic inputs received by tectal neurons, we recorded evoked excitatory postsynaptic currents in response to optic chiasm stimulation. We found that the amplitude of the early, monosynaptic component of evoked responses (the average current between 5 and 15 ms after the shock; Figure 3B) differed across experimental groups (Figure 3A; F(5,161)=3.2, p=0.009; see "Methods" for a description of the linear model we used). Both Sync and Async multisensory groups had larger early synaptic currents than the Control group (Tukey p=0.03 and 0.04; Cohen d=0.92 and 0.73 on log-transformed data respectively). The amplitude of late synaptic currents–produced by recurrent network activation (15-145 ms after the stimulus)–did not differ across groups.

>>>>

It is often assumed, in fields as diverse as connectomics and machine learning, that the main difference between functional and dysfunctional neural networks lies in their connectivity \citep{takemura2014,hildebrand2017,bassett2017network,reimann2017}. Biological neurons, however, are tuned in ways that go well beyond adjusting one “strength” value per synapse: cells within the same network often demonstrate variability of activation thresholds \citep{kole2012}, production of bursts \citep{popovic2011}, inactivation by strong inputs \citep{bianchi2012}, and more. This diversity of tuning relies on coordinated changes of many parameters across a complex, multivariate landscape \citep{oleary2013}, as neuronal phenotypes are shaped by sensory experiences, and adjusted by modulatory inputs \citep{evans2015}. The dysregulation of intrinsic plasticity affects network dynamics \citep{tien2018}, and can lead to a loss of function \citep{marcelin2009}. And yet, with the obvious exception of oscillatory networks \citep{marder2011,picton2018control}, for many brain areas it is still unclear whether variation of intrinsic phenotypes serves as a defining aspect of network topology and architecture, or whether it is just a consequence of transfer function normalization \citep{titley2017}.

The optic tectum of the \textit{Xenopus} tadpole is an ideal model for exploring these questions: it is a highly malleable distributed network of about $10^4$ neurons \citep{pratt2013}, involved in stimulus discrimination and sensorimotor transformations \citep{dong2009,khakhalin2014}. In development, tectal neurons acquire diverse intrinsic and synaptic phenotypes that are also shaped by sensory experiences \citep{xu2011,ciarleglio2015}. Circuits in the tectum can learn and reproduce the temporal dynamics of inputs to which they were exposed \citep{pratt2008}: a property that could in principle be achieved through synaptic changes alone \citep{lukovsevivcius2009}, but which is more likely to involve intrinsic temporal tuning \citep{narayanan2008,beatty2014}. Finally, tectal neurons exhibit strong Na channel inactivation, which seems to play a role in collision detection \citep{jang2016}, and is one of the targets for intrinsic plasticity \citep{bianchi2012}.

In this study, we asked two specific questions about the properties of intrinsic plasticity in tectal networks. First, we asked whether intrinsic plasticity is limited to changes in excitability, or whether it is more nuanced and can differentially adjust neuronal responsiveness to inputs with different dynamics. Second, we checked whether changes in intrinsic properties in the tectum are a homeostatic response to each cell's history of synaptic activation, or if they are independent of synaptic properties. In our previous large-scale census of tectal cells \citep{ciarleglio2015}, we observed no interaction between intrinsic and synaptic phenotypes, and despite an extensive search, we did not detect signs of temporal tuning (ibid,  figures 2, 4). However, we propose that the standard current-clamp protocols used in previous studies \citep{pratt2007,hamodi2014} were not adequate to detect important changes in the function of voltage-gated channels (see Discussion).

Here we show that the intrinsic plasticity of tectal neurons goes beyond changes in average spikiness and supports temporal selectivity that can be reshaped by sensory experience. Furthermore, we show that the tuning of intrinsic properties is coordinated with the duration of synaptic inputs received by each cell. These results rely on two methodological innovations. First, instead of using current injections, we employed the dynamic clamp technique, which allowed a more realistic simulation of synaptic conductances \citep{prinz2004}. Unlike more common voltage and current clamp techniques, in dynamic clamp the electric current injected into the cell is dynamically adjusted, based on a predefined formula that depends on cell membrane potential and time. Second, instead of relying on one type of sensory stimulation, we used five different stimulation protocols and compared their effects on intrinsic tuning. The use of different sensory modalities also gave us insight into an unrelated, but equally intriguing question of multisensory integration in the brain \citep{deeg2009,felch2016,truszkowski2017}, as for the first time we were able to look at tectal network retuning in response to multisensory stimuli in freely behaving tadpoles.

@@

All analysis scripts and summary data for every cell can be found at:  \url{https://github.com/khakhalin/Dynamic-clamp-2018}

@@

Our first question was whether our stimulation protocols caused changes in intrinsic excitability of tectal neurons. From previous studies, we knew that in tadpoles exposed to four hours of LED flashes, tectal neurons became more excitable \citep{aizenman2003,ciarleglio2015}. However, the stimuli we used in the present study were weaker, and similar to those used in behavioral experiments \citep{khakhalin2014,james2015,truszkowski2017}. We presented a checkerboard pattern that inverted once a second either instantaneously (Figure 1C left, dubbed “Flash”), or with a slow transition over the course of a second (old black squares shrank to white, while new black squares grew from old white squares; Figure 1C right; dubbed “Looming”; see Methods).

\begin{figure*}
\includegraphics[width=\linewidth]{fig1.pdf}
\caption{
Overview of experimental design and summary of dynamic clamp results. (\textbf{A}). Positions of tectal neurons that were recorded. (\textbf{B}). Sample data from a dynamic clamp experiment. Bottom row: the dynamics of conductances G(t) of four different durations simulated by the dynamic clamp system. Middle row: the currents I(t) dynamically injected into a cell based on conductances of 4 different durations and 3 different amplitudes. Top row: resulting voltage traces V(t) that were recorded and analyzed. (\textbf{C}). A schematic of visual conditioning in “Flash” (left) and “Looming” (right) groups. (\textbf{D}). The number of spikes produced by all neurons in all experiments, split by input peak conductance, and plotted against conductance duration. Black lines show respective averages. (\textbf{E}). A summary of data from (d), presented as averages and 95\% confidence intervals.}
\end{figure*}

After conditioning, we excised the brain, obtained whole cell patch clamp recordings from neurons in the tectum (Figure 1A), and counted spikes produced during dynamic clamp in response to simulated synaptic conductances of different durations and amplitudes (Figure 1B). We used conductances of 4 different durations (100, 200, 500, and 1000 ms), and 3 different amplitudes (peak conductances of 0.2, 0.5, and 1.0 nS), matching the range of synaptic currents observed in tectal circuits \textit{in vivo} (see Methods). Contrary to our expectations, in tadpoles that were exposed to instantaneous checkerboard inversions (flashes), tectal neurons on average became not more, but less spiky, and generated 0.4$\pm$0.4 spikes, across all types of dynamic clamp inputs, compared to 0.9$\pm$1.0 spikes in control (Figure 1D, left column; F(1,677)=30.4, p=5e-8, n cells=28, 29. Here and below, F-values are reported for a multivariate fixed effects analysis of variance with selected interactions, where cell id is used as a repeated measures factor; see Methods for a detailed description). Neurons exposed to looming transitions also spiked less than control neurons (0.6$\pm$0.4 spikes; F(1,629)=9.6; p=0.002; n=28, 25), but more than those exposed to instantaneous "flashes" (F(1,641)=3.7; p=4e-6; n=29, 25), which was likewise unexpected. In the tadpole tectum, looming stimuli are known to elicit stronger responses compared to instantaneous inversions \citep{khakhalin2014}, yet the suppression yielded by looming stimuli was weaker.

We then mapped the amplitude tuning of neurons (also known as amplitude transfer function, or gain), by looking at how an increase in transmembrane conductance translated into increased spike output. Compared to control, neurons from animals exposed to visual stimuli had a flatter amplitude tuning curve (Fig 1D,E), and did not increase their spiking as fast in response to larger conductances (F(1,677)=15.9, p=8e-5; and F(1,629)=8.3, p=0.004 for flash- and looming transitions respectively). The flattening of response curves seemed slightly more pronounced for flashes than for looming, but this difference was not significant (F(1,641)=3.5; p=0.06).

@@

Across input conductances of different lengths (100, 200, 500, and 1000 ms), neurons exposed to flashes and looming stimuli responded differently than in control (F(1, 677)=25.1; p=7e-7; and F(1,629)=12.0; p=6e-4 respectively). In control neurons, longer inputs typically evoked stronger spiking, whereas neurons from stimulated animals had flatter tuning curves, with a plateau, or even a decrease in spike number for longer conductance injections (Figure 1E). In essence, while control neurons “preferred” longer inputs, stimulated neurons developed a preference for shorter synaptic inputs, and this change was more pronounced in neurons exposed to flashes than in those exposed to looming stimuli (F(1,641)=7.5; p=0.006). This suggests that the change in overall intrinsic excitability, and the change in temporal tuning, follow two different kinds of logic. The overall excitability is homeostatic, as neurons became less excitable in response to stronger stimulation. The temporal retuning however can be better described as “adaptive”, as neurons exposed to shorter stimuli (flashes) became relatively \textit{more} responsive to shorter stimuli, and less responsive to longer stimuli, which is opposite of what one would expect for a fully homeostatic retuning. We chose to call this type of plasticity "adaptive", as presumably it means that after exposure to faster stimuli, neurons become more equipped to process faster patterns of activation \citep{stemmler1999information}.

@@

To test this question, we exposed tadpoles to four hours of behaviorally salient “click” sounds \citep{james2015,truszkowski2017}, provided at the same frequency (every second) as visual stimuli in the first set of experiments. We found (Fig 1D, E) that exposure to startle-inducing sounds (group “Sound”) did not lead to significant changes in either average spikiness (0.8$\pm$0.7 spikes; F(1,689)=1.7; p=0.2; n=28, 30), amplitude transfer function (F(1,689)=2.0; p=0.2), or temporal tuning curve (F(1,689)=2.0; p=0.2). This suggests that acoustic stimuli did not strongly activate tectal circuits during conditioning, despite being more behaviorally salient (for our stimuli, at the onset of stimulation, acoustic clicks evoked startle responses in about 50-80\% of cases, compared to 5-10\% for checkerboard inversions \citep{james2015,truszkowski2017}). As one possible explanation, auditory and mechanosensory inputs may have different cellular or subcellular targets in the tectum \citep{bollmann2009}, or they may differentially recruit tectal inhibitory circuits \citep{liu2016,hamodi2016}.

We then combined visual and acoustic stimuli in two different ways and looked at the effects of multisensory stimulation on the intrinsic properties of tectal neurons. For some animals, we synchronized the instantaneous checkerboard inversions (flashes) with sound clicks (dubbed “Sync”), while for others we staggered visual and acoustic stimuli by half a period (500 ms; dubbed “Async”). We found (Fig 1D, E) that, after four hours of multisensory stimulation, tectal neurons were more excitable than after visual stimulation alone (0.6$\pm$0.4 spikes, F(1,689)=11.2, p=8e-4; and 0.7$\pm$0.6 spikes, F(1,665)=41.7, p=2e-10, for sync and async respectively, across all testing conditions). Compared to the “Flash” group, the tuning curves in multisensory groups were less flat, with a stronger effect in the Async group (for amplitude tuning: F(1,689)=1.3, p=0.3, and F(1,665)=2.4, p=0.02 in Sync and Async groups respectively; for temporal tuning: F(1,689)=9.3, p=0.002, and F(1,665)=22.8, p=2e-6 respectively). This suggests that on their own sound clicks had little effect on tectal excitability, but when added to visual flashes, sound clicks negated effects of retuning that visual stimulation would have had (Fig 2C).

\begin{figure*}[!t]
\includegraphics[width=\linewidth]{fig2.pdf}
\caption{
Quantification of changes in temporal tuning in response to sensory experience. (\textbf{A}). An illustration of how “Temporal tuning” and “Amplitude tuning” values were calculated. For the temporal tuning measure, the value of zero corresponds to linear dependency (blue line), positive values - to an accelerating, supralinear curve (red), and negative values - to a plateau-shaped curve (purple, yellow). For amplitude tuning, higher values correspond to faster increase in spiking with increased conductance. (\textbf{B}). Amplitude tuning of neurons across different experimental groups (stars show t-test p$<$0.05 compared to control). (\textbf{C}). Temporal tuning and average spikiness of neurons in different experimental groups; in each plot all neurons across all groups are shown in gray, while neurons from one target group are shown in color; means are shown as black dots; ellipses represent 95\% normal confidence regions. Two outliers (top right corner) are brought within the axes limits. (\textbf{D}). Same data as in (C), shown as averages for each group, with 95\% confidence intervals. Black arrows show the effects of sound clicks, when they were added to control, and when they were added to “Flashes” to form two types of multisensory stimuli.}
\end{figure*}

@@

To visualize and interpret differences in neuronal tuning, we quantified each of the three aspects of intrinsic tuning (average spikiness, amplitude tuning, and temporal tuning) with one value per neuron (see Methods). We used the mean number of spikes across all conditions as the measure of "spikiness"; the linear slope of the number of spikes as a function of conductance amplitude as the measure of "amplitude tuning", and the quadratic term of the curvilinear regression for the number of spikes as a function of input duration as the value characterising the "temporal tuning" or adaptation index of each neuron (Figure 2A). The numerical values of these "tuning coefficients" are not easily interpretable, but they capture the character of tuning curves for each neuron (Figure 2A). All three parameters differed across experimental groups: F(5,160)=3.1, p=0.01 for average spikiness (Figure 2C; see Methods for model description); F(5,160)=4.8, p=4e-4 for amplitude tuning (Figure 2B); and F(5,160)=3.6, p=4e-3 for temporal tuning (Figure 2C).

@@

Describing neuronal tuning with only a few variables allowed us to compare cell-to-cell variability of tuning in different experimental groups. We found that this variability decreased as neurons were modulated away from the baseline (Bartlett test p=2e-9 for amplitude tuning, p=1e-9 for temporal tuning; Figure 2D). Groups that were significantly different in average values also had different variances (F-test with p$<$0.05), such as Control vs. Flash and Control vs. Looming for both amplitude and temporal tuning. This expands on a finding in our previous study \citep{ciarleglio2015} that prolonged patterned stimulation reduces diversity of tuning profiles in the network, reshaping them according to the spatiotemporal characteristics of the stimulus, with stronger stimuli having stronger effects on tuning diversity.

@@

To see whether prolonged sensory stimulation affected synaptic inputs received by tectal neurons, we recorded evoked excitatory postsynaptic currents in response to optic chiasm stimulation. We found that the amplitude of the early, monosynaptic component of evoked responses (the average current between 5 and 15 ms after the shock; Figure 3B) differed across experimental groups (Figure 3A; F(5,161)=3.2, p=0.009; see Methods for a description of the linear model we used). Both Sync and Async multisensory groups had larger early synaptic currents than the Control group (Tukey p=0.03 and 0.04; Cohen d=0.92 and 0.73 on log-transformed data respectively). The amplitude of late synaptic currents–produced by recurrent network activation (15-145 ms after the stimulus)–did not differ across groups.
